Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-22 cover_letter.tex
Up title page

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic beliefs, and preferences for market justice in
Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with greater support for redistributive policies; however,
individuals with strong meritocratic convictions are more likely to
legitimize existing disparities. The study also considers the influence
of major social movements during this period, which appear to have
reshaped public discourse and perceptions of fairness. These findings
contribute to a deeper understanding of how beliefs about justice and
equity evolve in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{1.15}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
assets (\protect\hyperlink{ref-chancel_world_2022}{Chancel et al.,
2022}). This context of economic disparity has sparked renewed interest
in studying not only the objective aspects of inequality, such as income
and access to resources, but also its subjective dimensions, including
perceptions, beliefs, and associated attitudes
(\protect\hyperlink{ref-janmaat_subjective_2013}{Janmaat, 2013}). This
has led, among other reasons, to widespread concern in the social
sciences about the political delegitimization that economic inequality
can produce (\protect\hyperlink{ref-castillo_perception_2022}{Castillo
et al., 2022}). Understanding how people perceive inequality is crucial,
as these perceptions can influence how societies comprehend and justify
the distribution of goods and services.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society
(\protect\hyperlink{ref-akyelken_urban_2020}{Akyelken, 2020}).
Regardless of their measurement, various studies have shown that this
perception often underestimates the gap between the rich and the poor,
which could have implications for attitudes toward the distribution of
goods and services
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022}; \protect\hyperlink{ref-schroder_income_2017}{Schröder, 2017}).
Moreover, in recent years, there has been a discussion about the
subjective and objective aspects of inequality, demonstrating
inconsistencies between the two
(\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). This tension
is relevant because advancing the study of perceptions could help
understand how perceived inequality affects the distribution of
resources within a society (Becker, 2020).

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022};
\protect\hyperlink{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et
al., 2020},
\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{2018}). Research
shows that greater perceived inequality fosters stronger preferences for
redistribution, regardless of actual inequality levels
(\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
al., 2019};
\protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). A less explored
area in this regard concerns market justice, which denotes the degree to
which individuals consider just that social goods and services (e.g.,
health, pensions, education) are allocated according to individual
contribution, competition, and ability to pay
(\protect\hyperlink{ref-kluegel_legitimation_1999}{Kluegel et al.,
1999}; \protect\hyperlink{ref-lane_market_1986}{Lane, 1986};
\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}). Market justice
attitudes reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\protect\hyperlink{ref-lane_market_1986}{Lane,
1986}). Indeed, Lindh (\protect\hyperlink{ref-lindh_public_2015}{2015})
shows that market justice preferences are stronger in countries with
higher private spending on services than in those with more
comprehensive welfare systems. From this perspective, perceptions of
economic inequality may be key, as lower perceived gaps can reinforce
acceptance of free-market systems and justify existing inequalities.

Meritocracy posits that inequality can be legitimized through
distributive criteria such as effort and talent
(\protect\hyperlink{ref-davis_principles_2001}{Davis \& Moore, 2001};
\protect\hyperlink{ref-young_rise_1962}{Young, 1962}). Previous studies
have demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\protect\hyperlink{ref-mijs_paradox_2021}{Mijs,
2021}; \protect\hyperlink{ref-wilson_role_2003}{Wilson, 2003}) and also
to justify more inequality as these beliefs are associated with
attitudes that legitimize status differences
(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}). In
a context like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020}), these
beliefs can play a crucial role in the acceptance of social
inequalities. Incorporating the perception of meritocracy in this
literature allows for an understanding of how attitudes toward
inequality are shaped not only by the perception of economic gaps, but
also by beliefs about how these gaps are justified. Furthermore, these
beliefs are often consolidated from an early age, reinforced by
institutions promoting values such as effort and individual skills to
climb socially
(\protect\hyperlink{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \protect\hyperlink{ref-reynolds_perceptions_2014}{Reynolds \&
Xian, 2014}).

The perception of meritocracy and the perception of economic inequality
would interact intricately in shaping preferences for market justice. On
the one hand, a stronger perception of merit as a criterion for
distribution could minimize the perception of economic gaps, thus
justifying an unequal system
(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
2012}; \protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021}). On the
other hand, the perception of inequality could moderate the impact of
meritocracy, as a higher perception of economic gaps might challenge the
idea that these are solely based on merit.

The primary objective of this study is to analyze the interplay between
perceptions of inequality and meritocracy and their joint influence on
preferences for market justice in Chile from 2016 to 2023, using
longitudinal data from the ELSOC survey. This interaction is expected to
provide a more comprehensive explanation of attitudes toward market
justice in a country characterized by high inequality and strong
free-market influence
(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020};
\protect\hyperlink{ref-flores_top_2020}{Flores et al., 2020}).
Additionally, this analysis seeks to elucidate how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services. Recognizing the temporal
dimension in shaping market justice preferences is essential, given that
such preferences are not static but are influenced by historical and
contextual factors that challenge or reaffirm prevailing social norms.
In this regard, variations in perceived economic inequality and
meritocracy over time can affect how individuals endorse market-based
approaches. Consequently, public opinion may shift from supporting
market justice to embracing redistributive policies aimed at mitigating
inequalities.

\hypertarget{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}{%
\section{Theoretical views on market justice, inequality perception, and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}}

\hypertarget{the-justification-of-market-inequality}{%
\subsection{The justification of market
inequality}\label{the-justification-of-market-inequality}}

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane
(\protect\hyperlink{ref-lane_market_1986}{1986}), who makes a contrast
between market justice and political justice. The author defines market
justice as a system of ``earned deserts,'' whereby individuals are seen
as deserving of a determined distributive outcome due to their effort
and skills. In contrast, political justice emphasizes principles of
equality and need, which are often represented by the welfare state
action through social policies. An important remark is that the
principles of market justice rely on the assumption that markets are
neutral and self-regulating arenas, where individuals are treated fairly
because they face the same formal rules of engagement
(\protect\hyperlink{ref-lane_market_1986}{Lane, 1986}). Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. In this way, market justice
provides a moral justification for inequality by framing it as a
necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\protect\hyperlink{ref-kluegel_beliefs_1981}{1981}) who assessed the
normative foundations that explain public support for economic
inequality. Over time, this approach has been extended beyond income to
include other market-mediated outcomes, such as education, healthcare,
and/or pensions. For example, Von Dem Knesebeck et al.
(\protect\hyperlink{ref-vondemknesebeck_are_2016}{2016}) and Immergut
and Schneider (\protect\hyperlink{ref-immergut_it_2020}{2020}) examine
whether citizens consider it fair that individuals with higher incomes
can access better healthcare, while Lee and Stacey
(\protect\hyperlink{ref-lee_fairness_2023}{2023}) apply a similar method
in the context of education in Australia. These studies usually rely on
a survey item asking respondents to evaluate the fairness of
income-based access to welfare services, allowing for comparing justice
perceptions across different contexts. Similarly, comparative studies
have also considered these indicators to study the support for
market-based distribution in welfare systems
(\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015};
\protect\hyperlink{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\protect\hyperlink{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles
(\protect\hyperlink{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\protect\hyperlink{ref-svallfors_political_2007}{Svallfors, 2007}). This
tendency reflects not only material self-interest but also a broader
moral economy, in which winners of the market system internalize
justifications for the status quo. At the same time, political ideology
also plays a role ---such as economically conservative values --- where
more right-wing individuals show higher support for meritocracy and more
skepticism towards redistribution. This is particularly salient in
countries with more restricted public provision of social services. For
example, Castillo et al.
(\protect\hyperlink{ref-castillo_socialization_2024}{2024}) in Chile
and, Lee and Stacey (\protect\hyperlink{ref-lee_fairness_2023}{2023}) in
Australia show that those with right-leaning individuals are more
supportive of market-based distribution of welfare. Beyond individual
characteristics, country-level institutions also play a central role. In
liberal welfare regimes like those of the United States or the United
Kingdom, market justice preferences are more widespread, while in
coordinated or social-democratic regimes---such as Sweden or
Germany---citizens are generally more critical of market-based
inequalities (\protect\hyperlink{ref-immergut_it_2020}{Immergut \&
Schneider, 2020}; \protect\hyperlink{ref-lindh_public_2015}{Lindh,
2015}). These findings suggest that market justice is also shaped
contextually, which is represented by the development of welfare
institutions, in line with the policy-feedbacks literature
(\protect\hyperlink{ref-campbell_policy_2012}{Campbell, 2012};
\protect\hyperlink{ref-pierson_when_1993}{Pierson, 1993}).

\hypertarget{perception-of-inequality}{%
\subsection{Perception of inequality}\label{perception-of-inequality}}

Perceptions of inequality have been associated with attitudes about
market justice (\protect\hyperlink{ref-kluegel_social_1995a}{Kluegel et
al., 1995}; \protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}).
Research indicates that lower perceived inequality can reinforce support
for market-based distributive arrangements by suggesting that the system
is fair and that outcomes reflect effort and ability
(\protect\hyperlink{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when
inequality is perceived as excessive or structurally determined,
individuals could question the legitimacy of market justice and become
more supportive of redistributive policies
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022};
\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
al., 2019}).

Perceptions of inequality refer to individuals' subjective evaluations
of the extent, causes, and consequences of income and wealth
disparities. Unlike objective measures such as the Gini index, perceived
inequality captures how individuals make sense of distributive
hierarchies in their everyday lives, shaped by reference groups, social
comparisons, and information environments
(\protect\hyperlink{ref-garcia-castro_perceiving_2020}{García-Castro et
al., 2020};
\protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}; \protect\hyperlink{ref-mijs_stratified_2016}{Mijs,
2016}). Scholars have proposed multiple dimensions of perceived
inequality, including its magnitude (how significant are the gaps),
vertical structure (between which groups), the trend over time
(increasing or decreasing), and legitimacy (whether it is just or not)
(\protect\hyperlink{ref-engelhardt_what_2018}{Engelhardt \& Wagener,
2018};
\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
al., 2019}). These dimensions encompass both cognitive and normative
aspects of perceptions of inequality and can vary across societies and
social groups, depending on exposure, ideology, and personal experience
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022};
\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
et al., 2018}).

Among the different approaches to measuring perceived inequality, one of
the most widely used is the estimation of wage gaps between occupational
extremes, such as between a CEO and a manual worker. This type of item
provides a concrete frame that respondents can relate to more easily
than abstract questions about national income distribution
(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
2012}; \protect\hyperlink{ref-easterbrook_social_2021}{Easterbrook,
2021}; \protect\hyperlink{ref-willis_legitimacy_2015}{Willis et al.,
2015}). While it enables the estimation of inequality using simple
heuristics, this method is not without challenges. For instance, people
often lack reliable knowledge about the earnings of those at the top of
the income ladder, which leads to high variability in responses and the
use of biased mental shortcuts
(\protect\hyperlink{ref-knell_perceptions_2020}{Knell \& Stix, 2020}).
Despite this, perceived wage gaps are strong predictors of political
attitudes
(\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
et al., 2018}; \protect\hyperlink{ref-pedersen_attitudes_2019}{Pedersen
\& Mutz, 2019}), making them a valuable tool for understanding public
responses to economic disparities.

Another relevant critique of this approach lies in its conflation of
different psychological constructs. Many surveys assess perceived
inequality through Likert-type items that ask respondents to agree or
disagree with statements, such as ``income differences are too large,''
which captures general concern or discomfort rather than a specific
perception (\protect\hyperlink{ref-Castillo2011}{Castillo, 2011};
\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
al., 2019}). These items mix cognitive estimations with affective
evaluations, complicating the interpretation of what respondents
perceive versus what they morally reject. As a result, the conceptual
clarity between perceived inequality and inequality aversion remains
blurred in many empirical studies. To address this limitation, recent
work has emphasized the need to distinguish between absolute and
comparative measures, as well as between ideal and actual estimates of
economic gaps (\protect\hyperlink{ref-auspurg_why_2017}{Auspurg et al.,
2017};
\protect\hyperlink{ref-garcia-sanchez_creencias_2022}{García-Sánchez \&
De Carvalho, 2022}).

\hypertarget{perception-of-meritocracy}{%
\subsection{Perception of meritocracy}\label{perception-of-meritocracy}}

Meritocracy constitutes a central ideological framework for legitimizing
different types of social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\protect\hyperlink{ref-young_rise_1962}{1962}), the term was meant to
critique a system in which merit-based stratification becomes a new form
of inequality. However, over time, meritocracy has been widely supported
in many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
\protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}). From a
sociological perspective, the belief in meritocracy is more than a
cognitive assessment; it reflects a moral lens through which individuals
interpret inequality. People who believe that success results from hard
work and talent are more likely to view social and economic disparities
as legitimate (\protect\hyperlink{ref-batruch_belief_2023}{Batruch et
al., 2023}; \protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et
al., 2012}). Conversely, if they see outcomes as driven by luck, social
origin, or systemic barriers, inequality is more likely to be perceived
as unjust. This distinction becomes crucial in societies with persistent
structural inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

We adopt a multidimensional perspective on meritocracy, distinguishing
between two key dimensions: effort-based and talent-based perceptions.
This distinction is essential, as it captures different pathways through
which individuals justify inequality. Effort-based meritocracy
emphasizes hard work and perseverance as the basis for social rewards,
aligning closely with cultural narratives of personal responsibility. A
talent-based meritocracy, by contrast, emphasizes intelligence and
innate abilities, which are often perceived as less malleable and more
unequally distributed. Both dimensions have been shown to correlate with
acceptance of inequality, but they may carry distinct implications for
how inequality is justified in specific domains
(\protect\hyperlink{ref-castillo_multidimensional_2023}{Castillo et al.,
2023}). The relevance of this distinction is supported by recent
studies, which show that individuals respond differently to these
dimensions. For instance, perceptions that effort is rewarded in society
are more strongly associated with positive evaluations of fairness and
acceptance of unequal outcomes
(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}).
This may be because effort is seen as a controllable and morally
virtuous trait, whereas talent is often perceived as a natural
advantage. Consequently, effort-based meritocracy is likely more potent
in legitimizing inequality, particularly in neoliberal contexts.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle

\setstretch{1.15}
\section{Authors Identification}\label{authors-identification}

Juan Carlos Castillo, Department of Sociology, Universidad de Chile,
Santiago de Chile\footnote{Corresponding author (juancastillov@uchile.cl
  -
  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile).}.\\
ORCID: 0000-0003-1265-7854

Andreas Laffert, Institute of Sociology, Pontificia Universidad Católica
de Chile.\\
ORCID: 0000-0002-9008-2454

Kevin Carrasco, Center for Social Conflict and Cohesion Studies.\\
ORCID: 0000-0002-9348-8361

Julio Iturra, Bremen International Graduate School of Social Sciences,
Bremen, Germany. ORCID: 0000-0003-1993-3479

\section{Exclusive Submission
Statement}\label{exclusive-submission-statement}

The authors confirm that this article has been submitted exclusively to
this journal.

\section{Acknowledgments}\label{acknowledgments}

The authors recognize the support of the National Research and
Development Agency (ANID) through the grant from the National Fund for
Scientific and Technological Development (FONDECYT) N° 1210847; from the
Center for Conflict and Social Cohesion Studies (COES) grant Financing
Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and
Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS)
NCS2022\_046.

\section{Authors' Contributions}\label{authors-contributions}

Author roles were classified using the Contributor Role Taxonomy
\href{https://credit.niso.org/}{CRediT} as follows: Juan Carlos
Castillo: Conceptualization, Supervision, Methodology, Writing - review
and editing; Andreas Laffert: Conceptualization, Methodology, Formal
analysis, Software, Visualization, Writing - review and editing; Kevin
Carrasco: Methodology, Formal analysis, Writing - review and editing.
Julio Iturra: Conceptualization, Writing - review and editing.

\section{Conflict of Interest}\label{conflict-of-interest}

The authors declare no conflict of interest.

\section{Funding}\label{funding}

This research received funding from ANID/FONDECYT grant 1210847 and the
Center for Conflict and Social Cohesion Studies (COES) through
ANID/FONDAP grant 1513009.

\section{Data Avaibility Statement}\label{data-avaibility-statement}

The original data presented in the study are openly available in
\url{https://dataverse.harvard.edu/api/access/datafile/10797987}

Access to the data, code, and analysis presented in the study are openly
available in
\url{https://github.com/justicia-distributiva/market-justice-panel-elsoc}.
